Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2016-05-02 cikm2016/intro.tex
More text trimming
Make more academic
I hope!

The SQL-99 standard defines three temporal scalar types: \texttt{DATE}, \texttt{TIMESTAMP} and \texttt{TIME}, which can be further qualified as either \texttt{WITH} or \texttt{WITHOUT} \texttt{TIME ZONE} (the \texttt{WITHOUT} form being more common.) They are typically implemented as fixed-point types containing an offset from some epoch (\eg Julian Days.) This makes them compact to store using column store compression techniques such as those in C-Store~\cite{Stonebraker:2005} and MonetDB/X100~\cite{Zukowski:2006}, and further allows some temporal operations to be implemented very efficiently using simple arithmetic. Thus from the RDBMS perspective, representing dates in scalar form provides  benefits for users, both in terms of analytic richness and query performance.

@@

\subsection{Parsing Dates}
A common version of the date parsing problem is converting columns of integers in the form \texttt{yyyyMMdd} to date scalars. \Naive\ users often solve this problem by converting the integer to a string and performing some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach has a number of problems because string operations are notoriously slow compared to scalar operations (typically 10-100x slower in modern RDBMSes). Default parsing of date formats is locale-dependent, and may not work when the parsing expression is shared across an international organization (\eg between the US and European offices). The parsing code is often hard to understand and maintain because it uses a verbose, general-purpose string-handling syntax instead of a specialized domain language. Further, there is no single data format.

@@

Several RDBMSes (\eg MySQL, Oracle and Postgres) provide row-level functions for parsing formatted dates, but our analysis of user data showed a 15\% syntax error rate using these functions. And even with perfect syntax, the user still has to interrupt her flow to learn a formatting syntax -- one that she will likely forget after this specific problem has been solved.

@@

Our solution was to develop two algorithms for automatically deriving the format string from the user's data with over $95\%$ parsing accuracy. Both algorithms are built based on classical machine learning algorithms that learn from pattern recognition to make data-driven predictions. We developed two algorithms because we were unaware of any previous work on this problem and we needed to be able to cross-validate our results on a large testing corpus. With both approaches, the user can simply specify that the column is a date, and the visualization system can respond quickly and accurately enough to avoid interrupting the user's flow.

@@

Specifically, our contributions include:
\begin{itemize}
\setlength\itemsep{0em}
\item By analyzing an online corpus, we provide evidence that practical date parsing requires the ability to recognize hundreds of formats;
\item We show how to extend prior work on Minimum Descriptive Length structure extraction to generate a freely available date format domain language with over 95\% accuracy;
\item We describe a second Natural Language Processing technique for generating the same date format domain language with similar accuracy. We describe how the basic algorithm is extended to support grammar variants and constraints unique to date formats. The parsing algorithm is extended to compute an overall dominant pattern over a data column;
\item We provide evidence that the development of multiple, independent parsing algorithms provides an effective means of cross-validation on large corpora;
\item We describe some limitations of this domain language that would improve its utility.
\end{itemize}
%% vidya: don't think we need this.
%% richard: It's kind of traditional in the DB community, but we can just leave it commented out - and put it back if they complain!
%\subsection{Organization}
%The rest of this paper is organized as follows: The next section provides background information for the problem space. The following two sections describe the two different algorithms, one using Minimum Descriptive Length and the other using Natural Language Processing. In section V, we evaluate the algorithms on a corpus of 30K columns, both by sampling the outputs manually and then by using the algorithms to validate each other. We then discuss related work in section VI, future work in section VII and conclude in section VIII.

>>>>

The SQL-99 standard defines three temporal scalar types: \texttt{DATE}, \texttt{TIMESTAMP} and \texttt{TIME}, which can be further qualified as either \texttt{WITH} or \texttt{WITHOUT} \texttt{TIME ZONE} (the \texttt{WITHOUT} form being more common.) They are typically implemented as fixed-point types containing an offset from some epoch (\eg Julian Days.) This makes them compact to store using columnar compression techniques such as those in C-Store~\cite{Stonebraker:2005} and MonetDB/X100~\cite{Zukowski:2006}, and further allows some temporal operations to be implemented efficiently using simple arithmetic. 
Thus from the RDBMS perspective, representing dates in scalar form provides benefits for users, both in terms of analytic richness and query performance.

@@

\subsection{Parsing Dates}
A common version of the date parsing problem is converting columns of integers in the form \texttt{yyyyMMdd} to date scalars. \Naive\ users often solve this problem by converting the integer to a string and performing some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach has a number of problems because string operations are notoriously slow compared to scalar operations (typically 10-100x slower in modern RDBMSes). Default parsing of date formats is locale-dependent, and may not work when the parsing expression is shared across an international organization (\eg between the US and European offices). The parsing code is often hard to understand and maintain because it uses a verbose, general-purpose string-handling syntax instead of a specialized domain language.

@@

Several RDBMSes (\eg MySQL, Oracle and Postgres) provide row-level functions for parsing formatted dates, but a simple review of user attempts to use this functionality in published analyses showed a $15\%$ syntax error rate when using these functions unassisted. And even with perfect syntax, the user still has to interrupt her flow to learn a formatting syntax -- one that she will likely forget after this specific problem has been solved.

@@

Our solution was to develop two algorithms for automatically deriving the format string from the user's data with over $95\%$ parsing accuracy. Both algorithms are built based on classical machine learning algorithms that learn from pattern recognition to make data-driven predictions. We developed two algorithms because we were unaware of any previous work on this problem and we needed to be able to cross-validate our results on a large test corpus. With both approaches, the user can simply specify that the column is a date, and the visualization system can respond quickly and accurately enough to avoid interrupting the user's flow.

@@

\begin{itemize}
\setlength\itemsep{0em}
\item By analyzing an online corpus, we provide evidence that practical date parsing requires the ability to recognize hundreds of formats;
\item We show how to extend prior work on Minimum Descriptive Length structure extraction to generate a freely available date format domain language with over 95\% accuracy;
\item We describe a second Natural Language Processing technique for generating the same date format domain language with similar accuracy. We describe how the basic algorithm is extended to support grammar variants and constraints unique to date formats. The parsing algorithm is extended to compute an overall dominant pattern over a data column;
\item We provide evidence that the development of multiple, independent parsing algorithms provides an effective means of cross-validation on large corpora;
\item We describe some limitations of this domain language that would improve its utility.
\end{itemize}
%% vidya: don't think we need this.
%% richard: It's kind of traditional in the DB community, but we can just leave it commented out - and put it back if they complain!
%\subsection{Organization}
%The rest of this paper is organized as follows: The next section provides background information for the problem space. The following two sections describe the two different algorithms, one using Minimum Descriptive Length and the other using Natural Language Processing. In section V, we evaluate the algorithms on a corpus of 30K columns, both by sampling the outputs manually and then by using the algorithms to validate each other. We then discuss related work in section VI, future work in section VII and conclude in section VIII.
